---
title: "p8105_hw3_pt2632"
output: 
  github_document:
  html_document:
    toc: TRUE
    float_toc: TRUE
date: "2023-10-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r import-lib}
library(tidyverse)
library(p8105.datasets)
```
# Problem 1
```{r}
# load dataset
data("instacart")
```

## Data Description

There are `r ncol(instacart)` variables and `r nrow(instacart)` in this datasest. 

structure of the data, describing some key variables, and giving illstrative examples of observations


## Answers to Questions 
1. How many aisles are there, and which aisles are the most items ordered from?

There are `r length(unique(pull(instacart, aisle_id)))` aisles in total. 
The aisle with id `r instacart |> group_by(aisle_id) |> summarize(n_orders = n()) |> mutate(max_order = max(n_orders)) |> filter(n_orders == max_order) |> pull(aisle_id)` are the aisle with most order items.

2. Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r}
instacart |> 
  # group operation
  group_by(aisle_id) |> 
  # calculate order count of each aisle
  summarize(n_orders = n()) |>
  # select the aisle with more than 10000 ordered items
  filter(n_orders > 10000) |>
  # arrange aisle
  arrange(aisle_id) |>
  # scatter point plot of "order counts" over "aisle id"
  ggplot(aes(aisle_id, n_orders)) + geom_point(alpha = .4) +
  labs(
    title = "Order count in aisle", 
    x = "aisle id", 
    y = "order count") +
  theme(plot.title = element_text(hjust = 0.5))
```
Comments: 


3. Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r}
instacart |>
  # filter entry from specific aisle
  filter(aisle == "baking ingredients" | 
         aisle == "dog food care" | 
         aisle == "packaged vegetables fruits") |>
  # group operation
  group_by(aisle, product_id) |>
  # calculate the count of ordered item of specific aisle
  summarize(order_count = n(), .groups = "drop_last") |>
  # calculate the rank of order count in decreasing order
  mutate(rnk = min_rank(desc(order_count))) |>
  # filter the 3 most popular products of specific aisle 
  filter(rnk < 4) |>
  # arrange data
  arrange(aisle, rnk) |> 
  # knit as table
  knitr::kable(digits = 2)
```

Comments: 



4. Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}
instacart |> 
  # filter the specific product
  filter(product_name == "Pink Lady Apples" |
         product_name == "Coffee Ice Cream") |>
  # group operation
  group_by(product_name, order_dow) |>
  # calculate the mean order hour of each product in specific day in a week
  summarize(mean_order_hour = mean(order_hour_of_day), .groups = "drop_last") |>
  # transform rows to columns
  pivot_wider(names_from = order_dow, values_from = mean_order_hour, names_prefix = "day_") |>
  # knit as table
  knitr::kable(digits = 2)
```


Comments: 



# Problem 2
```{r}
# load dataset
data("brfss_smart2010")
```

## Data Cleaning
```{r}
# define level order of response
ordered_response <- c("Poor", "Fair", "Good", "Very good", "Excellent")
cleaned_brfss <- brfss_smart2010 |> 
  # tidy colnames
  janitor::clean_names() |>
  # select topic "Overall Health"
  filter(topic == "Overall Health",
         # select response from “Excellent” to “Poor”
         response %in% ordered_response) |>
  # Order response from "Poor" to "Excellent"
  mutate(response = factor(response, levels = ordered_response))
```
Description:



## Answers to Questions
1. In 2002, which states were observed at 7 or more locations? What about in 2010?

In 2002, states including `r cleaned_brfss |> filter(year == 2002) |> select(locationabbr, locationdesc) |> distinct() |> group_by(locationabbr) |> summarize(n_location = n()) |> filter(n_location >= 7)` were observed at 7 or more locations.

In 2010, states including `r cleaned_brfss |> filter(year == 2010) |> select(locationabbr, locationdesc) |> distinct() |> group_by(locationabbr) |> summarize(n_location = n()) |> filter(n_location >= 7)` were observed at 7 or more locations.


2. Construct a dataset that is limited to `Excellent` responses, and contains, year, state, and a variable that averages the `data_value` across locations within a state. 
Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the `geom_line` geometry and `group` aesthetic will help).
```{r}
cleaned_brfss |>
  # filter entrys with "Excellent" response
  filter(response == "Excellent") |>
  # group operation
  group_by(year, locationabbr) |>
  # calculate the mean of data_value of each group
  summarize(mean_val = mean(data_value), .groups = "drop_last") |>
  # line plot
  ggplot(aes(x = year, y = mean_val)) + 
  geom_line(aes(color = locationabbr, group = locationabbr), alpha = .5) +
  labs(
    title = "Average value over time for states", 
    x = "year", 
    y = "mean value") +
  theme(plot.title = element_text(hjust = 0.5))
```
Comments: 



3. Make a two-panel plot showing, for the years 2006, and 2010, distribution of `data_value` for responses (“Poor” to “Excellent”) among locations in NY State.
```{r}
cleaned_brfss |>
  # filter entry from specific year
  filter(year == 2006 | year == 2010) |>
  # density plot for each response type
  ggplot(aes(x = data_value, fill = response)) +
  geom_density(alpha = .5) +
  # plot panels
  facet_grid(year ~ .) +
  labs(
    title = "Distribution of data_value", 
    x = "data_value") +
  theme(plot.title = element_text(hjust = 0.5))
```
Comments: 




# Problem 3
```{r}
# load data
acc <- tibble(read.csv("./datasets/nhanes_accel.csv"))
covar <- tibble(read.csv("./datasets/nhanes_covar.csv", skip = 4))
```

## Data Manipulation
Load, tidy, merge, and otherwise organize the data sets. 
Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).
```{r}
data <- 
  # merge 2 dataset
  left_join(covar, acc, by = join_by(SEQN == SEQN)) |>
  # filter entry with age >= 21 and non-NA demographic data
  filter(age >= 21 & !is.na(sex) & !is.na(age) & !is.na(BMI) & !is.na(education)) |>
  # change variable class (numeric to factor)
  mutate(sex = factor(
    case_match(sex, 1 ~ "male", 2 ~ "female"), 
    levels = c("male", "female")), 
         education = factor(
           case_match(education, 
                      1 ~ "Less than high school",
                      2 ~ "High school equivalent", 
                      3 ~ "More than high school"), 
           levels = c("Less than high school", "High school equivalent", "More than high school"))
)
```
Description: 



## Answers to Questions
1. Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.
```{r}
data |> 
  # group operation
  group_by(sex, education) |>
  # calculate the count of entries in each group
  summarize(count = n(), .groups = "drop_last") |> 
  # knit as table
  knitr::kable(digits = 0)
```
Comments:


```{r}
data |>
  # density plot for different education levels and sex
  ggplot(aes(x = age)) + geom_density(aes(fill = sex), alpha = .4) + facet_grid(education ~ .) +
  labs(
    title = "Age distribution of different sex and education background", 
    x = "age") +
  theme(plot.title = element_text(hjust = 0.5))
```

2. Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.
```{r}
data |> 
  # transform columns to rows
  pivot_longer(cols = starts_with("min"), names_to = "min_label", values_to = "minites") |>
  # group operation
  group_by(SEQN, sex, education, age) |>
  # calculate the total activity time of each participant
  summarize(total_activity_time = sum(minites), .groups = "drop_last") |>
  # point plot and smooth plot
  ggplot(aes(x = age, y = total_activity_time)) + 
  geom_point(alpha = .4, color = "tomato") + geom_smooth() +
  facet_grid(sex ~ education) +
  labs(
    title = "Total activity time over age", 
    x = "age", 
    y = "mean value") +
  theme(plot.title = element_text(hjust = 0.5))
```

Comments:


3. Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

```{r}
data |> 
  # transform columns to rows 
  pivot_longer(cols = starts_with("min"), names_to = "min_label", values_to = "minites") |>
  # parse time label
  mutate(time_tick = parse_number(min_label)) |>
  # plot time course of 24-hour activity
  ggplot(aes(x = time_tick, y = minites, color = sex)) +
  geom_smooth() +
  facet_grid(education ~ .) +
  labs(
    title = "24-hour activity course", 
    x = "time", 
    y = "acticity") +
  theme(plot.title = element_text(hjust = 0.5))
```
Description:

Conclusion:


